Novel lattice Boltzmann method for simulation of strongly shear thinning viscoelastic fluids

### 2024-02-29 NovelLBM.tex
remove lines

	\begin{center}
		$^1$ Biofluid Simulation and Modeling – Theoretische Physik VI, University of Bayreuth, Bayreuth, Germany
		$^2$ FMZ
	\end{center}
	\vspace{\fill}
	\begin{center}
		%TODO abstact
	\end{center}
	%TODO keywords

@@

For a simple shear flow in the $xy$-plane of a PTT fluid with $ \xi =0$, an analytical expression for $ \eta $ and $N_1$ can be derived as detailed in appendix~\ref{app:semianalyticalPoiseuille}:
	\begin{equation}
		\eta\left(\dot{\gamma}\right) = \frac{\eta_\text{p}}{\exp\left[0.5W_0\left(4\epsilon \mathrm{Wi}^2\right)\right]} + \eta_\text{s} \label{eq:shearEta}
	\end{equation}
	\begin{equation}
		N_1 = \frac{\eta_\text{p}}{2\epsilon\lambda}W_0\left(4\epsilon \mathrm{Wi}^2\right) \label{eq:shearN}
	\end{equation}
where $\mathrm{Wi} = \lambda\dot{\gamma}$ is the Weissenberg number and $W_0$ is the Lambert $W$-function.
Fitting these equations to the rheological data for alginate and methyl cellulose leads to satisfactory agreement as shown in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc}.
In the fit, the solvent viscosity was fixed at $\eta_\text{s} = \SI{1}{\milli\pascal\second}$ as the viscosity of the PBS buffer. 
%\SG{ give ref }\RK{PBS is 99\% water. If you really want a citation tho we could use doi: 10.1063/1.4823586} \SG{ ok, let's leave it out. }
%At very high shear rates, the contained polymers are completely stretched and the viscosity asymptotically approaches the solvent value $ \eta _ \mathrm{s}$.
The obtained parameters are given in table~\ref{tab:PTTparameters}.
	\begin{table}[H]
		\centering
		\begin{tabular}{c|c|c|c|c}
			Fluid & $\eta_\text{p}/\unit{\milli\pascal\second}$ & $\lambda/\unit{\milli\second}$ & $\epsilon$ & $R_\eta$ \\
			\hline
			alginate \SI{4}{\percent} & \num{48.2\pm0.4e3} & \num{343 \pm 4} & \num{0.545 \pm 0.010} & \num{4.8e4} \\
			methyl c. \SI{0.49}{\percent} & \num{18.7\pm0.4} & \num{0.344\pm0.003} & \num{0.270\pm0.003} & \num{18.7} \\
			methyl c. \SI{0.59}{\percent} & \num{32.5\pm0.7} & \num{0.433\pm0.004} & \num{0.365\pm0.004} & \num{32.5} \\
			methyl c. \SI{0.83}{\percent} & \num{81\pm2} & \num{0.714\pm0.009} & \num{0.496\pm0.006} & \num{81}
		\end{tabular}
		\caption{Parameters of the PTT model obtained by fitting to the rheological measurements in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc}.
		\SG{ add a column with $R_ \eta $ }}\RK{would be identical to the content of the $\eta_p$ column, also $R_\eta$ would be used before introduced} \SG{ Stimmt, aber ich will es auf einen Blick sichtbar machen }
		\label{tab:PTTparameters}
	\end{table}

@@

\begin{table}[H]
	\centering
		\begin{tabular}{l|p{3cm}|p{1.8cm}|p{1.8cm}|p{1.8cm}|p{2cm}}
			Author & Material & $\eta_0/\unit{\pascal\second}$ & $\eta_\infty/\unit{\pascal\second}$ & $R_\eta$ & extracted from \\
			\hline
			Amorim\cite{amorim2021} & pre-crosslinked alginate &  $>\num{1e4}$ & $<\num{3e-1}$ & $>\num{3e4}$ & Fig. 3b \\
			\rowcolor{secondaryBackground}
			Pössl\cite{possl2021} & blend & \num{2e2} & \num{5e-1} & \num{4e2} & Figure 4 \\
%			Paxton\cite{paxton2017} &Nivea Creme\n Alginate& 1e5\n5e4 & <1\n<5e1 & >1e5\n>1e3 & Fig. 4b \\
			Paxton\cite{paxton2017} & Alginate& \num{5e4} & $<\num{5e1}$ & $>\num{1e3}$ & Fig. 4b \\
			\rowcolor{secondaryBackground}
			Jalaal\cite{jalaal2017} & pluronic F-127 & $>\num{1e3}$ & $<\num{1e-1}$ & $>\num{1e4}$ & FIG. 6. (e) \\
		\end{tabular}
	\caption{Compilation of literature data on typical bioinks showing that a very viscosity ratio $ R_ \eta $ is a generic feature of these liquids.
	\SG{ cross-check these values }\RK{Done and formatted}
	}
	\label{tab:bioinks}
\end{table}

@@

\section{Existing LBM methods}

\SG{ make sure that all papers cited in the introduction as ''3D'' are discussed here as well }\RK{dellar is missing}
\SG{ and vice versa: make sure that all papers discussed here are also cited in the introduction }\RK{gupta2015\_2 not in intro} \SG{ dann setz einfach die Zitate so rein, wie du sie fuer passend haeltst. }\RK{done}
Since the early days of LBM, variants for viscoelastic fluids have been developed.
Keeping in mind our target geometries, we focus here on three-dimensional formulations.
The first approaches avoided an explicit modeling of a viscoelastic constitutive equation and simply modified the LB collision operator to obtain a basic viscoelastic fluid obeying Jeffery's model \cite{lallemand2003}. This can also be archived by the introduction of a second local stress tensor \cite{dellar2014}.
More advanced approaches include the work of Onishi \textit{et al.} \cite{onishi2005, onishi2006} who considered a non-shear-thinning Oldroyd-B model using a microscopic Fokker-Planck equation for the polymer dumbbells.
The work of \cite{wang2020} solved the advection part of the model using a second LBM running in parallel to the Navier-Stokes LBM.
The same technique was used by \cite{malaspinas2010} and \cite{ma2020} for the Oldroyd-B and FENE-P models.
Later, hybrid schemes coupling an LB solver for the solvent to a finite-difference \cite{gupta2015_2, gupta2015, gupta2016} or finite-volume \cite{kuron2021} scheme for the polymer stress (or, alternatively, the polymer conformation) tensor were introduced.
To improve numerical stability, some authors propose the inclusion of an artificial diffusivity \cite{malaspinas2010,ma2020,wang2020,wang2021}, which introduces a small error compared to the analytical solution of the constitutive equation.
Others perform special decompositions of the stress tensor\cite{dzanic2022}.
Coupling of the resulting polymer stress to the Navier-Stokes LBM can be achieved via the inclusion of point forces or via a modification of the equilibrium populations.
Both methods have recently been compared in 2D \cite{dzanic2022_2}.

@@

To include the polymer stress into the LBM, we use the stress coupling scheme according to Dzanic \textit{et al.} \cite{dzanic2022_2, gupta2015, onishi2006}.
For this, ... \SG{ more details }\todo{SRT collision operator as implemented, advection as implemented, the whole as implemented appendix that was deleted}
To handle inflow/outflow boundaries, the polymer conformation tensor is specified on the boundary and advected using the corner transport upwind scheme.

@@

The key difficulty for simulating many realistic viscoelastic fluids is the imbalance between the viscous stress of the solvent and the polymer stress.
While the former is handled efficiently and reliably by the LBM streaming-collision scheme, the latter enters the LB equations as an additional source term.
The relative magnitude of the two terms is given by the viscosity ratio $R_\eta$ as defined in equation~\ref{eq:R_eta} which for many situations is of the order of $10^2-10^3$ (see figures~\ref{fig:fit_alginate}, \ref{tab:bioinks} and table~\ref{tab:bioinks}).
The key idea of our scheme is to artificially increase the LBM viscous stress while at the same time reducing the polymer stress by the same amount.
Starting from equation~\ref{eq:totalStress}, we write
\begin{align}
	\ma{\sigma} = 2\eta_\text{s}\ma{D} + \ma{\tau}_\text{shuffle} + \ma{\tau} - \ma{\tau}_\text{shuffle}	\label{eq:shuffleStress}
\end{align}
with the transferred ("shuffled") stress
\begin{align}
	\ma{\tau}_\text{shuffle} = 2\alpha_\text{s}\eta_\text{p}\ma{D}
\end{align}
containing an arbitrary parameter $\alpha_\text{s}$. 
Finally, this results in
\begin{align}
	\ma{\sigma} = 2\left(\eta_\text{s} + \alpha_\text{s}\eta_\text{p}\right)\ma{D} + \left(\ma{\tau} - 2\alpha_\text{s}\eta_\text{p}\ma{D}\right).
\end{align}
In practice, the shuffling parameter $\alpha_\text{s}$ is chosen before the start of the simulation.
The LBM solver is run with $\eta_\text{s} + \alpha_\text{s}\eta_\text{p}$ as its viscosity.
The polymer stress determined by equation~\ref{eq:polymerStress} is reduced by the amount $\alpha_\text{s}\eta_\text{p}\ma{D}$ at every time step before it is coupled into the LBM.

@@

We start by simulating the alginate and \SI{0.49}{\percent} methyl cellulose solutions with a viscosity ratio $R_\eta = \num{48.2e3}$ and $R_\eta = \num{18.7}$, respectively, in 2D planar Poiseuille flow.
In physical units the channel has a height of \SI{20}{\micro\meter} which are discretized with 41 LBM nodes.
For the viscosity shuffling, we set $\alpha_\text{s} = 1$.
Along the flow direction, the channel is periodic, and the flow is driven by a pressure gradient of \SI{-1e7}{\pascal\per\meter} for alginate and  approximately \SI{-1.2e7}{\pascal\per\meter} for methyl cellulose.
The velocity field is initialized with zero.
For methyl cellulose, we simulate the full temporal evolution with the polymers starting in the unstretched configuration.
In the case of alginate, we speed up the convergence and start from a situation in which the polymers are pre-stretched \SG{ give details }\RK{initialized to the semianalytical solution} \SG{ schreib am besten eine Referenz auf die Gleichung hin, muesste ja im Appendix stehen. }\todo{do}.
After \num{2e9} LBM time steps for alginate and \num{5e6} LBM time steps for methyl cellulose corresponding to approximately \SI{1.6}{\milli\second} and \SI{2.4}{\second} of physical time, respectively, the simulation is stopped and the flow field $u(y)$ is compared to the semi-analytical solution.
%\RK{
%	Alg 2e9 1.6455647956111251 ms (reveals the sim to be a scam)
%	MC 5e6 10.065719379854297 ms
%}
As shown in figure~\ref{fig:2Dvel}, very good agreement is found.
For a quantitative assessment, we compute the L2 error as follows.
\begin{equation}
	e_\text{L2} = \sqrt{\frac{\sum_i\left[\vec{u_i} - \vec{u}\left(y_i\right)\right]^2}{\sum_i\vec{u}\left(y_i\right)^2}}
\end{equation}
which gives \num{9e-4} and \num{1e-3} for alginate and methyl cellulose, respectively.
The spatially resolved errors are shown in figure~\ref{fig:2DvelErr} in the appendix.
These conclusions remain unchanged even when varying the pressure gradient, and thus the flow velocity, over many orders of magnitude as shown in figure~\ref{fig:pressureErr}~(a) of the appendix.

@@

%Malaspinas: We use richard/master fe0fc04115b9b6ea6628bcc6cb6593d3f72ac0ea (2023-11-05)
Finally, we assess the quality of our algorithm by comparing to an existing study on viscoelastic LBM by Malapsinas \textit{et al.} \cite{malaspinas2010}.
The authors studied an Oldroyd-B fluid in a planar Poiseuille flow and compared the error in the simulated velocity profile to an exact analytical solution.
The Oldroyd-B model is a special case of the PTT model which is obtained by setting $\xi = \epsilon = 0$ in eq.~(\ref{eq:polymerStress}) and is thus easily tractable with our code.\RK{was simulated with a direct Oldroyd-B implementation, which I now see was unnecessary...} \SG{ hast du mal kurz geprueft, ob dasselbe rauskommt? }\RK{Nein}\todo{check}
In figure~\ref{fig:malaspinas2010}, we show the L2 error as function of lattice resolution for different values of $\mathrm{Wi}$ and $R_ \nu $ which is defined as \cite{malaspinas2010}
\begin{align}
	R_\nu = \frac{\nu_\text{s}}{\nu_\text{p}+\nu_\text{s}}
\end{align}
with the kinematic solvent and the polymer viscosity $\nu_\text{s}$ and $\nu_\text{p}$. 
We note that this can directly be transformed into the viscosity ratio used above as $\frac{1}{R_\nu}-1=R_\eta$.
Furthermore, in agreement with \cite{malaspinas2010}, also the L2 error is defined differently than above, namely:
	\begin{equation}\label{eq:l2error}
		E_\text{u} = \frac{1}{u_\text{max}}\sqrt{\frac{1}{N}\sum\abs{u_\text{simulation}-u_\text{theory}}^2}.
	\end{equation}
At this point, we mention that there is a typographic mistake in eq.~(61) of \cite{malaspinas2010} where the prefactor should be 8 instead of 4. \SG{ stimmt das so? }\RK{Ja}

@@

As can be seen in figure~\ref{fig:malaspinas2010}, the accuracy of our method does not depend on the Weissenberg Number and is generally somewhat better than the method of \cite{malaspinas2010}.
We note a certain dependence of the accuracy on $R_\nu$ which we consider nevertheless acceptable given the overall smallness of the error.
Finally, we find that the error of our method decreases with lattice resolution to the power of \num{-2}. \RK{They see diminishing returns we don't} \SG{ explain }\RK{The topmost line starts from the first green dot and also is a power of -2 line, the second point is already a little of and the third and fourth deviate more and more. One could try to do a line with a smaller power, however that does not quite fit the data either. This is particularly evident for the red points. It looks like Malaspinas has about power of -2 for small N but the slope decreases for higher N. This can be explained through other error terms (for example their artificial viscosity) becoming more relevant as the lattice resolution caused error decreases. As we do not have those our error continues to decrease as expected}

@@

The first application stems from bioprinting where the bioink is pushed through a conical nozzle \cite{possl2021, Poologasundarampillai.2021, oconnell2022}.
The nozzle radius is \SI{1.8}{\milli\meter} at the inlet and \SI{200}{\micro\meter} at the outlet.
The nozzle has a length of \SI{18}{\milli\meter}.
Due to the difference in inlet/outlet radii, the system cannot be modeled with periodic boundary conditions, and we use inflow/outflow boundary conditions as described in section~\ref{sec:Methods} instead.
The required values for velocity and polymer conformation are obtained from simulations of pipes of the respective radii and a fixed volume flow of $\approx\SI{3}{\micro\liter\per\second}$.
The resulting flow field can be seen in figure \ref{fig:Nozzle}~(a).
To quantify the total stress as a single number, we use the von Mises stress which is defined as \cite{Mueller.2023}:
\begin{align}
	\sigma_\text{vM} = \sqrt{\frac{1}{2}\left[\left(\sigma_1 - \sigma_2\right)^2 + \left(\sigma_2 - \sigma_3\right)^2 + \left(\sigma_3 - \sigma_1\right)^2\right]}.
\end{align}
It can be seen that the highest stresses arise at the nozzle exit in direct vicinity to the wall.
This means that even for fast flows, cells flowing close to the center experience less stress than cells flowing near the walls in agreement with earlier results in Newtonian fluids \cite{Mueller.2023}.
	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/Nozzle_v.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/Nozzle_sigma_vM.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/Nozzle_tau_12.pdf}
		\caption{Velocity profile (a), von Mises stress (b) and shear stress (c) for methyl cellulose in a conical nozzle as used for bioprinting. For the calculated viscosity see figure~\ref{fig:Nozzle/RT-DC_eta}.\RK{The black lines currently hide our resolution shall I remove them to be consistent with the RT-DC plots or do we prefer this?}}
		\label{fig:Nozzle}
	\end{figure}

@@

Our next example is the RT-DC geometry in which cells from a reservoir are squeezed through a narrow observation channel \cite{Toepfner.2018, Fregin:2019gx} with a square cross-section of 20x20 \unit{\micro\meter}.
For this simulation we use periodic boundary conditions and a constant body force corresponding to a pressure gradient of \SI{1e7}{\pascal}.
In figure~\ref{fig:RT-DC}, we show the velocity profile and the shear component of the stress which is the relevant component if a local shear viscosity is to be defined.
There are three qualitative differences compared to a purely viscous fluid.
Directly after the inlet we find a region of low stress.
Due to the fast advection, the polymers, despite being inside the constriction, have not reacted to the increased fluid stress yet.
Similarly, directly after the constriction, the high stress inside the observation channel persists over a certain distance that is needed for the polymer stress to decay.
\RK{There are three, proceeds to name two} \SG{ ? }\todo{do}

 	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/RT-DC_v.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/RT-DC_sigma_vM.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/RT-DC_tau.pdf}
		\caption{Velocity profile (a), von Mises stress (b) and shear stress (c) for methyl cellulose in a typical RT-DC geometry. For the calculated viscosity see figure~\ref{fig:Nozzle/RT-DC_eta}.
		\SG{ change to von Mises instead of $ \tau _{ 12 }$? }\RK{done something}}
		\label{fig:RT-DC}
	\end{figure}

>>>>

	\begin{center}
		$^1$ Biofluid Simulation and Modeling – Theoretische Physik VI, University of Bayreuth, Bayreuth, Germany\\
		$^2$ FMZ
	\end{center}
	\vspace{\fill}
	\begin{center}
		%TODO abstact
	\end{center}
	%TODO keywords

@@

For a simple shear flow in the $xy$-plane of a PTT fluid with $ \xi =0$, an analytical expression for $ \eta $ and $N_1$ can be derived as detailed in appendix~\ref{app:semianalyticalPoiseuille}:
	\begin{equation}
		\eta\left(\dot{\gamma}\right) = \frac{\eta_\text{p}}{\exp\left[0.5W_0\left(4\epsilon \mathrm{Wi}^2\right)\right]} + \eta_\text{s} \label{eq:shearEta}
	\end{equation}
	\begin{equation}
		N_1 = \frac{\eta_\text{p}}{2\epsilon\lambda}W_0\left(4\epsilon \mathrm{Wi}^2\right) \label{eq:shearN}
	\end{equation}
where $\mathrm{Wi} = \lambda\dot{\gamma}$ is the Weissenberg number and $W_0$ is the Lambert $W$-function.
Fitting these equations to the rheological data for alginate and methyl cellulose leads to satisfactory agreement as shown in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc}.
In the fit, the solvent viscosity was fixed at $\eta_\text{s} = \SI{1}{\milli\pascal\second}$ as the viscosity of the PBS buffer. 
%\SG{ give ref }\RK{PBS is 99\% water. If you really want a citation tho we could use doi: 10.1063/1.4823586} \SG{ ok, let's leave it out. }
%At very high shear rates, the contained polymers are completely stretched and the viscosity asymptotically approaches the solvent value $ \eta _ \mathrm{s}$.
The obtained parameters are given in table~\ref{tab:PTTparameters}.
	\begin{table}[H]
		\centering
		\begin{tabular}{c|c|c|c|c}
			Fluid & $\eta_\text{p}/\unit{\milli\pascal\second}$ & $\lambda/\unit{\milli\second}$ & $\epsilon$ & $R_\eta$ \\
			\hline
			alginate \SI{4}{\percent} & \num{48.2\pm0.4e3} & \num{343 \pm 4} & \num{0.545 \pm 0.010} & \num{4.8e4} \\
			methyl c. \SI{0.49}{\percent} & \num{18.7\pm0.4} & \num{0.344\pm0.003} & \num{0.270\pm0.003} & \num{18.7} \\
			methyl c. \SI{0.59}{\percent} & \num{32.5\pm0.7} & \num{0.433\pm0.004} & \num{0.365\pm0.004} & \num{32.5} \\
			methyl c. \SI{0.83}{\percent} & \num{81\pm2} & \num{0.714\pm0.009} & \num{0.496\pm0.006} & \num{81}
		\end{tabular}
		\caption{Parameters of the PTT model obtained by fitting to the rheological measurements in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc}}.
		\label{tab:PTTparameters}
	\end{table}

@@

\begin{table}[H]
	\centering
		\begin{tabular}{l|p{3cm}|p{1.8cm}|p{1.8cm}|p{1.8cm}|p{2cm}}
			Author & Material & $\eta_0/\unit{\pascal\second}$ & $\eta_\infty/\unit{\pascal\second}$ & $R_\eta$ & extracted from \\
			\hline
			Amorim\cite{amorim2021} & pre-crosslinked alginate &  $>\num{1e4}$ & $<\num{3e-1}$ & $>\num{3e4}$ & Fig. 3b \\
			\rowcolor{secondaryBackground}
			Pössl\cite{possl2021} & blend & \num{2e2} & \num{5e-1} & \num{4e2} & Figure 4 \\
%			Paxton\cite{paxton2017} &Nivea Creme\n Alginate& 1e5\n5e4 & <1\n<5e1 & >1e5\n>1e3 & Fig. 4b \\
			Paxton\cite{paxton2017} & Alginate& \num{5e4} & $<\num{5e1}$ & $>\num{1e3}$ & Fig. 4b \\
			\rowcolor{secondaryBackground}
			Jalaal\cite{jalaal2017} & pluronic F-127 & $>\num{1e3}$ & $<\num{1e-1}$ & $>\num{1e4}$ & FIG. 6. (e) \\
		\end{tabular}
	\caption{Compilation of literature data on typical bioinks showing that a very viscosity ratio $ R_ \eta $ is a generic feature of these liquids.
	}
	\label{tab:bioinks}
\end{table}

@@

\section{Existing LBM methods}
Since the early days of LBM, variants for viscoelastic fluids have been developed.
Keeping in mind our target geometries, we focus here on three-dimensional formulations.
The first approaches avoided an explicit modeling of a viscoelastic constitutive equation and simply modified the LB collision operator to obtain a basic viscoelastic fluid obeying Jeffery's model \cite{lallemand2003}. This can also be archived by the introduction of a second local stress tensor \cite{dellar2014}.
More advanced approaches include the work of Onishi \textit{et al.} \cite{onishi2005, onishi2006} who considered a non-shear-thinning Oldroyd-B model using a microscopic Fokker-Planck equation for the polymer dumbbells.
The work of \cite{wang2020} solved the advection part of the model using a second LBM running in parallel to the Navier-Stokes LBM.
The same technique was used by \cite{malaspinas2010} and \cite{ma2020} for the Oldroyd-B and FENE-P models.
Later, hybrid schemes coupling an LB solver for the solvent to a finite-difference \cite{gupta2015_2, gupta2015, gupta2016} or finite-volume \cite{kuron2021} scheme for the polymer stress (or, alternatively, the polymer conformation) tensor were introduced.
To improve numerical stability, some authors propose the inclusion of an artificial diffusivity \cite{malaspinas2010,ma2020,wang2020,wang2021}, which introduces a small error compared to the analytical solution of the constitutive equation.
Others perform special decompositions of the stress tensor\cite{dzanic2022}.
Coupling of the resulting polymer stress to the Navier-Stokes LBM can be achieved via the inclusion of point forces or via a modification of the equilibrium populations.
Both methods have recently been compared in 2D \cite{dzanic2022_2}.

@@

To include the polymer stress into the LBM, we use the stress coupling scheme according to Dzanic \textit{et al.} \cite{dzanic2022_2, gupta2015, onishi2006}.
Specifically, we implement the LBM collision as follows.
\begin{equation}
	f_i\left(\vec{x}+\vec{c_i}\Delta t, t+\Delta t\right) - f_i\left(\vec{x}, t\right) = -\frac{\Delta t}{\tau_\text{r}}\left[f_i\left(\vec{x}, t\right) - f_i^\text{eq}\left(\vec{x}, t\right) + B_i\right]
\end{equation}
Where $f_i$ are the populations, $f_i^\text{eq}$ are the equilibrium populations, $\vec{c_i}$ are the lattice velocities, $\Delta t$ is the time step, $\tau_\text{r}$ is the relaxation time and $B_i$ are the stress contributions defined as follows.
\begin{equation}
	B_i = w_i\left(\frac{\vec{c_i}\vec{c_i}}{2c_\text{s}^4}-\frac{\uT}{2c_\text{s}^2}\right):\ma{\tau}
\end{equation}
Where $w_i$ are the lattice weights, $c_\text{s}$ is the lattice speed of sound and $\ma{\tau}$ is the polymer stress tensor. The colon denotes the Frobenius inner product.
\SG{ more details }\RK{done}
To handle inflow/outflow boundaries, the polymer conformation tensor is specified on the boundary and advected using the corner transport upwind scheme.

@@

The key difficulty for simulating many realistic viscoelastic fluids is the imbalance between the viscous stress of the solvent and the polymer stress.
While the former is handled efficiently and reliably by the LBM streaming-collision scheme, the latter enters the LB equations as an additional source term.
The relative magnitude of the two terms is given by the viscosity ratio $R_\eta$ as defined in equation~\ref{eq:R_eta} which for many situations is of the order of $10^2-10^3$ (see figures~\ref{fig:fit_alginate}, \ref{tab:bioinks} and table~\ref{tab:bioinks}).
The key idea of our scheme is to artificially increase the LBM viscous stress while at the same time reducing the polymer stress by the same amount.
Starting from equation~\ref{eq:totalStress}, we write
\begin{align}
	\ma{\sigma} = 2\eta_\text{s}\ma{D} + \ma{\tau}_\text{shuffle} + \ma{\tau} - \ma{\tau}_\text{shuffle}	\label{eq:shuffleStress}
\end{align}
with the transferred ("shuffled") stress
\begin{align}
	\ma{\tau}_\text{shuffle} = 2\alpha_\text{s}\eta_\text{p}\ma{D}
\end{align}
containing an arbitrary parameter $\alpha_\text{s}$. 
Finally, this results in
\begin{align}
	\ma{\sigma} = 2\left(\eta_\text{s} + \alpha_\text{s}\eta_\text{p}\right)\ma{D} + \left(\ma{\tau} - 2\alpha_\text{s}\eta_\text{p}\ma{D}\right).
\end{align}
In practice, the shuffling parameter $\alpha_\text{s}$ is chosen before the start of the simulation.
The LBM solver is run with $\eta_\text{s} + \alpha_\text{s}\eta_\text{p}$ as its viscosity.
The polymer stress determined by equation~\ref{eq:polymerStress} is reduced by the amount $2\alpha_\text{s}\eta_\text{p}\ma{D}$ at every time step before it is coupled into the LBM.

@@

We start by simulating the alginate and \SI{0.49}{\percent} methyl cellulose solutions with a viscosity ratio $R_\eta = \num{48.2e3}$ and $R_\eta = \num{18.7}$, respectively, in 2D planar Poiseuille flow.
In physical units the channel has a height of \SI{20}{\micro\meter} which are discretized with 41 LBM nodes.
For the viscosity shuffling, we set $\alpha_\text{s} = 1$. The exact value of $\alpha_\text{s}$ does not matter as can be seen in figure~\ref{fig:ShuffleInPoiseuille}.
Along the flow direction, the channel is periodic, and the flow is driven by a pressure gradient of \SI{-1e7}{\pascal\per\meter} for alginate and  approximately \SI{-1.2e7}{\pascal\per\meter} for methyl cellulose.
The velocity field is initialized with zero.
For methyl cellulose, we simulate the full temporal evolution with the polymers starting in the unstretched configuration.
In the case of alginate, we speed up the convergence and start from a situation in which the polymers are pre-stretched. For this we calculate $\tau_{xx}$ and $\tau_{xy}$ using equations~(\ref{eq:tau_xx}) and (\ref{eq:PTTDerivation2}) using the $\dot{\gamma}$ calculated from equation~(\ref{eq:semiAnalyticGD}). The polymer stress is then initialized using these values. \SG{ give details }\RK{initialized to the semianalytical solution} \SG{ schreib am besten eine Referenz auf die Gleichung hin, muesste ja im Appendix stehen. }\RK{done}.
After \num{2e9} LBM time steps for alginate and \num{5e6} LBM time steps for methyl cellulose corresponding to approximately \SI{1.6}{\milli\second} and \SI{2.4}{\second} of physical time, respectively, the simulation is stopped and the flow field $u(y)$ is compared to the semi-analytical solution.
%\RK{
%	Alg 2e9 1.6455647956111251 ms (reveals the sim to be a scam)
%	MC 5e6 10.065719379854297 ms
%}
As shown in figure~\ref{fig:2Dvel}, very good agreement is found.
For a quantitative assessment, we compute the L2 error as follows.
\begin{equation}
	e_\text{L2} = \sqrt{\frac{\sum_i\left[\vec{u_i} - \vec{u}\left(y_i\right)\right]^2}{\sum_i\vec{u}\left(y_i\right)^2}}
\end{equation}
which gives \num{9e-4} and \num{1e-3} for alginate and methyl cellulose, respectively.
The spatially resolved errors are shown in figure~\ref{fig:2DvelErr} in the appendix.
These conclusions remain unchanged even when varying the pressure gradient, and thus the flow velocity, over many orders of magnitude as shown in figure~\ref{fig:pressureErr}~(a) of the appendix.

@@

%Malaspinas: We use richard/master fe0fc04115b9b6ea6628bcc6cb6593d3f72ac0ea (2023-11-05)
Finally, we assess the quality of our algorithm by comparing to an existing study on viscoelastic LBM by Malapsinas \textit{et al.} \cite{malaspinas2010}.
The authors studied an Oldroyd-B fluid in a planar Poiseuille flow and compared the error in the simulated velocity profile to an exact analytical solution.
The Oldroyd-B model is a special case of the PTT model which is obtained by setting $\xi = \epsilon = 0$ in eq.~(\ref{eq:polymerStress}) and is thus easily tractable with our code.\RK{was simulated with a direct Oldroyd-B implementation, which I now see was unnecessary...} \SG{ hast du mal kurz geprueft, ob dasselbe rauskommt? }\RK{Nein}\todo{check}
In figure~\ref{fig:malaspinas2010}, we show the L2 error as function of lattice resolution for different values of $\mathrm{Wi}$ and $R_ \nu $ which is defined as \cite{malaspinas2010}
\begin{align}
	R_\nu = \frac{\nu_\text{s}}{\nu_\text{p}+\nu_\text{s}}
\end{align}
with the kinematic solvent and the polymer viscosity $\nu_\text{s}$ and $\nu_\text{p}$. 
We note that this can directly be transformed into the viscosity ratio used above as $\frac{1}{R_\nu}-1=R_\eta$.
Furthermore, in agreement with \cite{malaspinas2010}, also the L2 error is defined differently than above, namely:
	\begin{equation}\label{eq:l2error}
		E_\text{u} = \frac{1}{u_\text{max}}\sqrt{\frac{1}{N}\sum\abs{u_\text{simulation}-u_\text{theory}}^2}.
	\end{equation}
At this point, we mention that there is a typographic mistake in eq.~(61) of \cite{malaspinas2010} where the prefactor should be 8 instead of 4.

@@

As can be seen in figure~\ref{fig:malaspinas2010}, the accuracy of our method does not depend on the Weissenberg Number and is generally somewhat better than the method of \cite{malaspinas2010}.
We note a certain dependence of the accuracy on $R_\nu$ which we consider nevertheless acceptable given the overall smallness of the error.
Finally, we find that the error of our method decreases with lattice resolution to the power of \num{-2}. The error found by Malaspinas \textit{et al.} has a similar decrease for small $N$, but sees less reduction at higher $N$.

@@

The first application stems from bioprinting where the bioink is pushed through a conical nozzle \cite{possl2021, Poologasundarampillai.2021, oconnell2022}.
The nozzle radius is \SI{1.8}{\milli\meter} at the inlet and \SI{200}{\micro\meter} at the outlet.
The nozzle has a length of \SI{18}{\milli\meter}.
Due to the difference in inlet/outlet radii, the system cannot be modeled with periodic boundary conditions, and we use inflow/outflow boundary conditions as described in section~\ref{sec:Methods} instead.
The required values for velocity and polymer conformation are obtained from simulations of pipes of the respective radii and a fixed volume flow of $\approx\SI{3}{\micro\liter\per\second}$.
The resulting flow field can be seen in figure \ref{fig:Nozzle}~(a).
To quantify the total stress as a single number, we use the von Mises stress which is defined as \cite{Mueller.2023}:
\begin{align}
	\sigma_\text{vM} = \sqrt{\frac{1}{2}\left[\left(\sigma_1 - \sigma_2\right)^2 + \left(\sigma_2 - \sigma_3\right)^2 + \left(\sigma_3 - \sigma_1\right)^2\right]}.
\end{align}
It can be seen in figure~\ref{fig:Nozzle}~(b) that the highest stresses arise at the nozzle exit in direct vicinity to the wall.
This means that even for fast flows, cells flowing close to the center experience less stress than cells flowing near the walls in agreement with earlier results in Newtonian fluids \cite{Mueller.2023}.
	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/Nozzle_v.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/Nozzle_sigma_vM.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/Nozzle_tau_12.pdf}
		\caption{Velocity profile (a), von Mises stress (b) and shear stress (c) for methyl cellulose in a conical nozzle as used for bioprinting. For the calculated viscosity see figure~\ref{fig:Nozzle/RT-DC_eta}.}
		\label{fig:Nozzle}
	\end{figure}

@@

Our next example is the RT-DC geometry in which cells from a reservoir are squeezed through a narrow observation channel \cite{Toepfner.2018, Fregin:2019gx} with a square cross-section of 20x20 \unit{\micro\meter}.
For this simulation we use periodic boundary conditions and a constant body force corresponding to a pressure gradient of \SI{1e7}{\pascal}.
In figure~\ref{fig:RT-DC}, we show the velocity profile, the von Mises stress and the shear component of the stress which is the relevant component if a local shear viscosity is to be defined.
There are three qualitative differences compared to a purely viscous fluid.
Directly after the inlet we find a region of low stress.
Due to the fast advection, the polymers, despite being inside the constriction, have not reacted to the increased fluid stress yet.
Similarly, directly after the constriction, the high stress inside the observation channel persists over a certain distance that is needed for the polymer stress to decay. If one would divide the stress in these two regions by the strain rate in an attempt to obtain a viscosity one would find a very low and a very high viscosity respectively. These can be outside the range of viscosities, that can be obtained through rheological measurements with the fluid. This effect is even more pronounced along the walls directly after the outlet. Here the viscosities obtained in this manner become negative. Due to the strain rate approaching $0$ at several locations within the simulation volume, this calculation often diverges. The clamped result can be seen in figure~\ref{fig:Nozzle/RT-DC_eta}. These discrepancies, especially the ones at the walls, show the breakdown of the viscosity interpretation in viscoelastic cases and signify the regions of interesting behavior. This also shows why viscoelastic simulations cannot always be replaced by purely viscous models and should be considered despite the high computational cost.
\SG{ decide whether we want to show a viscosity plot and remove or leave this part}\RK{done}
